Data Formatting

### 2021-02-19 scripts/data_format.Rmd
break up scripts so that all data prep is done in data_format and then sourced by other scripts

Packages
```{r}
library(tidyverse)
library(magrittr)
library(here)
```

@@

viral loads
```{r}
vl <- data %>% dplyr::select(vl.4:vl.42) %>% distinct() %>% pivot_longer(vl.4:vl.42)
vl$species_code <- as.double(gsub("vl.","",vl$name))
nrow(vl)
vl %<>% full_join(., names, by="species_code")
sum(vl$value>0)
```
17 total species; 12 species with non-zero viral loads

>>>>

Packages
```{r}
library(tidyverse)
library(magrittr)
library(vegan)

library(here)
```

@@

viral loads
```{r}
vl <- data %>% dplyr::select(vl.4:vl.42) %>% distinct() %>% pivot_longer(vl.4:vl.42)
vl$species_code <- as.double(gsub("vl.","",vl$name))
vl %<>% full_join(., names, by="species_code")
vl %<>% mutate(., ln_value = log(value)) %>% mutate(., log10_value = log10(value))
```
`r nrow(vl)` total species; `sum(vl$value>0)`species with non-zero viral loads

Transmission Model for two species - primarily used by fig1
```{r}
#define function contour(): inputs are parameters and output is values for grid search for R0
contour <- function(trans_1_min, trans_1_max, trans_3_min, trans_3_max, mort1, mort2, degr) {
  #expand data
  data <- expand_grid("trans1" = seq(from = trans_1_min, to = trans_1_max, by = 0.00001),
                      "trans2" = c(0.0001),
                      "trans3" = seq(from = trans_3_min, to = trans_3_max, by = 0.00001),
                      "shed1" = c(rep(1/2, length("trans1"))),
                      "shed2" = c(rep(1/2, length("trans1"))),
                      "reco1" = c(rep(1/10, length("trans1"))),
                      "reco2" = c(rep(1/10, length("trans1"))),
                      "mort1" = c(mort1), 
                      "mort2" = c(mort2), 
                      "degr" = c(degr), 
                      "birth1" = c(rep(5/3, length("trans1"))),
                      "birth2" = c(rep(5/3, length("trans1"))))

  #calculate disease-free equilibria for use in next-gen matrix
  data %<>% add_column(dfe1 = (data$birth1/data$mort1),dfe2 = (data$birth2/data$mort2)) 
  #make matrix
  mat <- matrix(1, nrow = 3, ncol = 3)

  m <- nrow(data)
  #calculate eigenvalue using next-gen matrix. max eigenvalue is R0
  for(n in 1:m){
    mat[1] = (data$trans1[n]*data$dfe1[n])/(data$reco1[n]+data$mort1[n])
    mat[2] = (data$trans1[n]*data$dfe2[n])/(data$reco2[n]+data$mort2[n])
    mat[3] = (data$shed1[n]/data$degr[n])
    mat[4] = (data$trans2[n]*data$dfe1[n])/(data$reco1[n]+data$mort1[n])
    mat[5] = (data$trans2[n]*data$dfe2[n])/(data$reco2[n]+data$mort2[n])
    mat[6] = (data$shed2[n]/data$degr[n])
    mat[7] = (data$trans3[n]*data$dfe1[n])/(data$reco1[n]+data$mort1[n])
    mat[8] = (data$trans3[n]*data$dfe2[n])/(data$reco2[n]+data$mort2[n])
    mat[9] = 0
    eigen <- max(eigen(mat)$values)
    data$eigen[n] <- eigen
  }

  data$span1 <- 1/data$mort1 #lifespan for species 1
  data$span2 <- 1/data$mort2 #lifespan for species 2
  data$half <- log(1/2)/-data$degr #viral half-life
  data$epid <- data$eigen > 1 #invasion potential (i.e. R0 > 1)
  data$prop1 <- data$trans1/data$trans2 #proportion of transmission of species 2 over species 1
  data$prop3 <- data$trans3/data$trans2 #proportion of environmental transmission over species 1
  data$prop_mort <- data$mort1/data$mort2 #relative mortality for species 1 and 2
  data$cap1 <- data$birth1/data$mort1 #carrying capacity for species 1
  data$cap2 <- data$birth2/data$mort2 #carrying capacity for species 2
  data$cap_prop <- data$cap1/data$cap2 #relative carrying capacities for species 1 and 2
  data$tot <- data$cap1 + data$cap2 #total abundance for species 1 and 2
  data$cc <- data$trans1*(data$cap1/data$tot) + data$trans2*(data$cap2/data$tot) #cc defined as transmission of a species multiplied by its relative abundance
  output <- data %>% dplyr::select(trans1, trans2, prop1, prop3, trans3, eigen, prop_mort, mort1, mort2, degr, half, cap1, cap2, cap_prop, tot, cc) %>% distinct()
  return(output)
}

reference <- contour(trans_1_min = 0.0001,trans_1_max = 0.001,trans_3_min = 0.0001,trans_3_max = 0.001,mort1 = 1/45,mort2 = 1/45,degr = 1/2)

composition <- contour(trans_1_min = 0.0001,trans_1_max = 0.001,trans_3_min = 0.0001,trans_3_max = 0.001,mort1 = 1/60,mort2 = 1/30,degr = 1/2)

size <- contour(trans_1_min = 0.0001,trans_1_max = 0.001,trans_3_min = 0.0001,trans_3_max = 0.001,mort1 = c(1/50),mort2 = c(1/50),degr = 1/2)

halflife <- contour(trans_1_min = 0.0001,trans_1_max = 0.001,trans_3_min = 0.0001,trans_3_max = 0.001,mort1 = c(1/40),mort2 = 1/40,degr = c(1/4))

combined <- contour(trans_1_min = 0.0001,trans_1_max = 0.001,trans_3_min = 0.0001,trans_3_max = 0.001,mort1 = c(1/66.6),mort2 = c(1/33.3),degr = c(1/4))

```

community matrix for PCA - primarily used by fig2
```{r}
#select for community variables
community_mat <- data %>% dplyr::select(WetAltID, Month.1, AB2:AB8, AB9, AB20:AB42) 
community_mat$Month.1 <- gsub("Month", "", community_mat$Month.1)
community_mat %<>% mutate(., siteID = paste(WetAltID, Month.1, sep = "_")) %>% distinct() %>% arrange(.,WetAltID) %>% dplyr::select(-WetAltID, -Month.1) %>% dplyr::select(siteID, AB2:AB42)
#rownames
community_mat %<>% column_to_rownames(., var = "siteID")

#select for environmental variables
env <- data %>% dplyr::select(WetAltID, Month.1, MeanAirT, MeanWaterTempPredC, DryingScore, CanopyCover, Area, Perimeter) %>% distinct() %>% arrange(.,WetAltID) 
env$Month.1 <- gsub("Month", "", env$Month.1)
#make siteID rowname
env$siteID <- paste(env$WetAltID, env$Month.1, sep = "_")
env %<>% dplyr::select(-WetAltID, -Month.1)
#remove duplicated rows (siteID is duplicated but env. variables are not... not sure why this is)
env <- env[-c(53, 90),]
env_mat <- env %>% column_to_rownames(., var = "siteID")

#replace NA's with zeros. Probably not the best option but will allow analysis to run for now
community_mat[is.na(community_mat)] <- 0
env_mat[is.na(env_mat)] <- 0

#pca on community matrix and gather scores for first two principal components
pca <- princomp(community_mat, scores = TRUE)
summary(pca)
scores <- as.data.frame(pca$scores)
scores %<>% dplyr::select(Comp.1, Comp.2) %>% rownames_to_column(., var = "siteID")

#plot pca
biplot(pca)

tmp <- data %>% dplyr::select(WetAltID, Month.1, cc, Month, AB2:AB42) %>% 
  dplyr::select(-ABAmb, -ABSal) %>%  mutate(., Size = rowSums(.[5:25])) %>% #get totals for community size
  dplyr::select(-AB2:-AB42) %>%
  mutate(., siteID = paste(WetAltID, Month.1, sep = "_")) %>%
  distinct()
tmp$siteID <- gsub("Month", "", tmp$siteID)

site_scores  <- inner_join(tmp, scores, by = "siteID")

#order months
site_scores$Month <- factor(site_scores$Month, levels = c("Feb", "Mar", "Apr", "May", "Jun", "Jul"))

#rank order components and plot in order
site_scores %<>% mutate(pc1Rank=dense_rank(Comp.1))
```

cc x evenness data - primarily used by fig3
```{r}
#Get species richness from matrix

community_mat <- data %>% dplyr::select(WetAltID, Month.1, AB2:AB8, AB9, AB20:AB42) %>%
  mutate(., siteID = paste(WetAltID, Month.1, sep = "_")) %>% distinct()

community_mat <- community_mat[order(community_mat$WetAltID, community_mat$Month.1),]

community_mat %<>% arrange(.,WetAltID) %>% dplyr::select(-WetAltID, -Month.1) %>% dplyr::select(siteID, AB2:AB42)

abundances <- community_mat

community_mat %<>% column_to_rownames(., var = "siteID")

presence_mat <- ifelse(community_mat>0, 1, 0)

siteID <- rownames(presence_mat)

richness_mat <- tibble()
for (i in 1:length(siteID)){
  richness_mat[i,1] <- siteID[i]
  richness_mat[i,2] <- sum(presence_mat[i,])
}
richness_mat %<>% rename(siteID = 1, richness = 2)

order <- richness_mat

#Get species richness for each site and community competence values

tmp <- data %>% dplyr::select(WetAltID, Month.1, cc, Prevalence,AB2:AB8, AB9, AB20:AB42) %>% 
  distinct() %>% 
  mutate(., siteID = paste(WetAltID, Month.1, sep = "_"))
tmp %<>% mutate(.,size = rowSums(.[5:25])) %>% select(-c(AB2:AB42))
tmp <- tmp[order(tmp$WetAltID, tmp$Month.1),]
tmp %<>% dplyr::select(-Month.1)

richness_cc <- full_join(tmp, richness_mat, by = "siteID")

#J = H'/ln(S)
#H' = Shannon diversity index
#S = Species Richness

h <- vegan::diversity(community_mat, index = "shannon")
h <- melt(as.matrix(h))
h %<>% rename(siteID = Var1, h = value) %>% dplyr::select(-Var2)

evenness <- richness_cc %>% mutate(lnS = log(richness)) %>% rowwise()

evenness %<>% full_join(h, evenness, by = "siteID")

evenness %<>% mutate(J = h/lnS) %>% rowwise()

#create new df for ordered sites to test lagged richness and evenness on prevalence
lag_evenness <- evenness[-c(4,14,47,90),]
lag_evenness$lag_richness <- c(0)
lag_evenness$lag_J <- c(0)
lag_evenness$lag_cc <- c(0)
lag_evenness$lag_size <- c(0)
for(n in 2:91){
  lag_evenness$lag_richness[n] <- lag_evenness$richness[n-1]
  lag_evenness$lag_J[n] <- lag_evenness$J[n-1]
  lag_evenness$lag_cc[n] <- lag_evenness$cc[n-1]
  lag_evenness$lag_size[n] <- lag_evenness$size[n-1]
}

#remove the first entry for each wetland to remove the carryover from the last wetland
lag_evenness %<>% group_by(WetAltID) %>% filter(duplicated(WetAltID) | n()==1)

#fit two lines to cc_evenness_prev plot
lag_evenness$high <- c(0)
lag_evenness$low <- c(0)
for (i in 1:nrow(lag_evenness)){
  if (lag_evenness$lag_J[i] > 0.6){
    lag_evenness$high[i] <- 1
    lag_evenness$low[i] <- 1
  }
  else if (lag_evenness$lag_cc[i] < 45000){
    lag_evenness$low[i] <- 1
  }
  else if (lag_evenness$lag_cc[i] > 45000){
    lag_evenness$high[i] <- 1
  }
  else {
    lag_evenness$high[i] <- 0
    lag_evenness$low[i] <- 0
  }
}
low_cc <- lag_evenness %>% filter(low==1)
cor.test(x=low_cc$lag_J, y=low_cc$lag_cc,method = "spearman")
high_cc <- lag_evenness %>% filter(high==1)
cor.test(x=high_cc$lag_J, y=high_cc$lag_cc,method = "spearman")
```

data for bar plots - primarily used by fig4
```{r}

#Get species richness from matrix

community_mat <- data %>% dplyr::select(WetAltID, Month.1, AB2:AB8, AB9, AB20:AB42) %>%
  mutate(., siteID = paste(WetAltID, Month.1, sep = "_")) %>% distinct() #select data and make siteID column

community_mat <- community_mat[order(community_mat$WetAltID, community_mat$Month.1),] #order by wetland and month

community_mat %<>% arrange(.,WetAltID) %>% dplyr::select(-WetAltID, -Month.1) %>% dplyr::select(siteID, AB2:AB42)

abundances <- community_mat #make abundances df

community_mat %<>% column_to_rownames(., var = "siteID") #make sitexspecies matrix

presence_mat <- ifelse(community_mat>0, 1, 0) #make pres-abs sitexspecies matrix

siteID <- rownames(presence_mat)

richness_mat <- tibble()
for (i in 1:length(siteID)){
  richness_mat[i,1] <- siteID[i]
  richness_mat[i,2] <- sum(presence_mat[i,])
}
richness_mat %<>% rename(siteID = 1, richness = 2) #make df with siteIDs and corresponding species richness

#Get species richness for each site and community competence values

tmp <- data %>% dplyr::select(WetAltID, Month.1, cc, Prevalence,AB2:AB8, AB9, AB20:AB42) %>% 
    distinct() %>% 
    mutate(., siteID = paste(WetAltID, Month.1, sep = "_"))
tmp %<>% mutate(.,size = rowSums(.[5:25])) %>% select(-c(AB2:AB42))
tmp <- tmp[order(tmp$WetAltID, tmp$Month.1),]
tmp %<>% dplyr::select(-Month.1)

richness_cc <- full_join(richness_mat, tmp, by = "siteID") #make df with siteIDs, richness, and cc

#J = H'/ln(S)
#H' = Shannon diversity index
#S = Species Richness

h <- vegan::diversity(community_mat, index = "shannon") #calculate shannons index for each site-month
h <- melt(as.matrix(h)) #melt into df
h %<>% rename(siteID = Var1, h = value) %>% dplyr::select(-Var2) #get df with shannons index for each site-month

evenness <- richness_cc %>% mutate(lnS = log(richness)) %>% rowwise() 
evenness %<>% full_join(h, evenness, by = "siteID")
evenness %<>% mutate(J = h/lnS) %>% rowwise() #calculate Pielou's J

#test for prevalence-richness relationship

evenness %>% ggplot(.,aes(x=richness, y=Prevalence))+
  geom_smooth(method="lm")
cor.test(evenness$richness,evenness$Prevalence,method="spearman")

#create new df for ordered sites to test lagged richness and evenness on prevalence
lag_evenness <- evenness[-c(4,14,47,90),] #remove site-months that are out of sequence
lag_evenness$lag_richness <- c(0)
lag_evenness$lag_J <- c(0)
lag_evenness$lag_cc <- c(0)
lag_evenness$lag_size <- c(0)
for(n in 2:91){ #lag values by one month
  lag_evenness$lag_richness[n] <- lag_evenness$richness[n-1]
  lag_evenness$lag_J[n] <- lag_evenness$J[n-1]
  lag_evenness$lag_cc[n] <- lag_evenness$cc[n-1]
  lag_evenness$lag_size[n] <- lag_evenness$size[n-1]
}
#remove the first entry for each wetland to remove the carryover from the last wetland
lag_evenness %<>% group_by(WetAltID) %>% filter(duplicated(WetAltID) | n()==1)
#lag_evenness %<>% select(siteID, lag_richness:lag_size)

lag_evenness %>% ggplot(.,aes(x=lag_richness, y=Prevalence))+
  geom_smooth(method="lm")
cor.test(lag_evenness$lag_richness,lag_evenness$Prevalence,method="spearman")

#barplots comparing community composition

evenness %<>% select(-WetAltID)
abundances %<>% left_join(.,evenness, by = "siteID")
lag_evenness %<>% select(siteID, lag_cc, lag_richness, lag_J, lag_size)
abundances %<>% inner_join(.,lag_evenness, by = "siteID")
abundances <- mutate(abundances, ABHigh = (AB9 + AB21 + AB26 + AB42)) 
abundances <- mutate(abundances, ABLow = (AB2 + AB3 + AB4 + AB5 + AB6 + AB8 + AB20 + AB24 + AB27 + AB28 + AB29 + AB31 + AB34 + AB35 + AB38 + AB39 + AB41))
abundances <- mutate(abundances, total = ABHigh + ABLow)
#final df for plots. includes data for abundance, prevalence, cc, diversity measure, lagged values

abundances$siteID <- reorder(abundances$siteID, -abundances$total) #order by total community size
#abundances$siteID <- reorder(abundances$siteID, -abundances$cc) #order by community competence
#abundances$siteID <- reorder(abundances$siteID, -abundances$J) #order by Pielou's J

```

correlations - primarily used by fig5
```{r}
prev_cc <- data %>% dplyr::select(WetAltID, Month.1, cc, Prevalence, Month, AB2:AB42, MeanWaterTempPredC)
prev_cc %<>% dplyr::select(-ABAmb, -ABSal) %>%  mutate(., total = rowSums(.[6:26])) #get totals for community size
prev_cc %<>% dplyr::select(-AB2:-AB42)
prev_cc %<>% distinct()

order <- prev_cc[order(prev_cc$WetAltID, prev_cc$Month.1),]

#this removes months out of sequence
order <- order[-c(4,14,47,90),]

order$lag_cc <- order$cc
order$lag_size <- order$total
order$lag_temp <- order$MeanWaterTempPredC

#create new column that includes previous month's value for cc
for(n in 2:nrow(order)){
  order$lag_cc[n] <- order$cc[n-1]
  order$lag_size[n] <- order$total[n-1]
  order$lag_temp[n] <- order$MeanWaterTempPredC[n-1]
}

#remove the first entry for each wetland to remove the carryover from the last wetland
clean <- order %>% group_by(WetAltID) %>% filter(duplicated(WetAltID) | n()==1)

```
